---
title: "Forecast daily bike rental demand using time series models"
date: "`r Sys.Date()`"
output: html_document
author: "Danyal Reyaz"
---

# About Data Analysis Report

This RMarkdown file contains the report of the data analysis done for the project on forecasting daily bike rental demand using time series models in R. It contains analysis such as data exploration, summary statistics and building the time series models. The final report was completed on `r date()`. 

**Data Description:**

This dataset contains the daily count of rental bike transactions between years 2011 and 2012 in Capital bikeshare system with the corresponding weather and seasonal information.

**Data Source:** https://archive.ics.uci.edu/ml/datasets/bike+sharing+dataset

**Relevant Paper:** 

Fanaee-T, Hadi, and Gama, Joao, 'Event labeling combining ensemble detectors and background knowledge', Progress in Artificial Intelligence (2013): pp. 1-15, Springer Berlin Heidelberg



**Task One: Load and explore the data**

**Load data and install packages**

```{r}
## Import required packages
# Install and load required packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(timetk, tidyverse, lubridate, ggplot2)

# Load the dataset
data("bike_sharing_daily")
bike_data <- bike_sharing_daily

# View the dataset
head(bike_data)
str(bike_data)
summary(bike_data)


```


**Describe and explore the data**

```{r}
# Convert the date column to Date type
bike_data$dteday <- as.Date(bike_data$dteday)

# Plot the rental counts over time
ggplot(bike_data, aes(x = dteday, y = cnt)) +
  geom_line() +
  labs(title = "Daily Bike Rentals", x = "Date", y = "Count")

# Correlation between temperature and total rentals
cor(bike_data$temp, bike_data$cnt)

# Mean and median temperatures for different seasons
bike_data %>%
  group_by(season) %>%
  summarize(mean_temp = mean(temp), median_temp = median(temp))

# Mean temperature, humidity, wind speed, and total rentals per month
bike_data %>%
  
  group_by(mnth) %>%
  summarize(mean_temp = mean(temp),
            mean_humidity = mean(hum),
            mean_windspeed = mean(windspeed),
            total_rentals = sum(cnt))

# Temperature association with bike rentals (registered vs. casual)
ggplot(bike_data, aes(x = temp)) +
  geom_point(aes(y = registered, color = "Registered")) +
  geom_point(aes(y = casual, color = "Casual")) +
  labs(title = "Temperature vs. Bike Rentals", x = "Normalized Temperature", y = "Count") +
  scale_color_manual(values = c("Registered" = "blue", "Casual" = "red"))
```



**Task Two: Create interactive time series plots**

```{r}
## Read about the timetk package
# ?timetk

# Create an interactive time series plot
bike_data %>%
  plot_time_series(.date_var = dteday, .value = cnt, .interactive = TRUE, .plotly_slider = TRUE, .color_var = year(dteday))

```




**Task Three: Smooth time series data**

```{r}
# Load additional required packages
pacman::p_load(forecast, zoo, TTR)

# Clean the time series data
bike_data_cleaned <- bike_data %>%
  mutate(cnt_clean = tsclean(ts(cnt, frequency = 365)))

# Plot cleaned data
ggplot(bike_data_cleaned, aes(x = dteday)) +
  geom_line(aes(y = cnt, color = "Original")) +
  geom_line(aes(y = cnt_clean, color = "Cleaned")) +
  labs(title = "Cleaned Daily Bike Rentals", x = "Date", y = "Count") +
  scale_color_manual(values = c("Original" = "blue", "Cleaned" = "red"))

# Apply Simple Moving Average (SMA)
bike_data_cleaned <- bike_data_cleaned %>%
  mutate(cnt_sma = SMA(cnt_clean, n = 10))

# Plot smoothed data
ggplot(bike_data_cleaned, aes(x = dteday)) +
  geom_line(aes(y = cnt_clean, color = "Cleaned")) +
  geom_line(aes(y = cnt_sma, color = "Smoothed (SMA)")) +
  labs(title = "Smoothed Daily Bike Rentals", x = "Date", y = "Count") +
  scale_color_manual(values = c("Cleaned" = "blue", "Smoothed (SMA)" = "red"))

# Apply Simple Exponential Smoothing
bike_ts <- ts(bike_data_cleaned$cnt_clean, frequency = 365)
fit_ets <- HoltWinters(bike_ts)

# Plot Exponential Smoothing
plot(fit_ets)
```



**Task Four: Decompse and access the stationarity of time series data**

```{r}
# Decompose the time series
decomp <- stl(bike_ts, s.window = "periodic")
plot(decomp)

# Check for stationarity using ADF test
library(tseries)
adf_test <- adf.test(bike_ts, alternative = "stationary")
adf_test$p.value

# If not stationary, apply differencing
bike_ts_diff <- diff(bike_ts)
adf_test_diff <- adf.test(bike_ts_diff, alternative = "stationary")
adf_test_diff$p.value

# Plot ACF and PACF for differenced data
acf(bike_ts_diff)
pacf(bike_ts_diff)
```



**Task Five: Fit and forecast time series data using ARIMA models**

```{r}
# Fit an ARIMA model
fit <- auto.arima(bike_ts, seasonal = TRUE)
summary(fit)

# Check residuals
checkresiduals(fit)

# Forecast future values
forecasted <- forecast(fit, h = 30)

# Plot the forecast
autoplot(forecasted) +
  labs(title = "Bike Rental Forecast for Next 30 Days", x = "Date", y = "Count")
```



**Task Six: Findings and Conclusions**

The analysis of daily bike rentals in the Capital Bikeshare system revealed several key insights:

1. **Seasonal Patterns**: Bike rentals exhibit clear seasonal patterns, with higher counts during warmer months and lower counts in colder months. This suggests that weather plays a significant role in bike rental behavior.

2. **Temperature Correlation**: There is a strong correlation between normalized temperature and the total count of bike rentals. Both casual and registered users show increased rentals with higher temperatures.

3. **Stationarity and ARIMA Model**: The time series data was not stationary initially, but after differencing, it became stationary. An ARIMA model was successfully fitted, and the forecast for the next 30 days provided reasonable predictions.

4. **Forecasting Accuracy**: The ARIMA model captured the overall trend and seasonality well, indicating that it can be a useful tool for predicting future bike rental demand.

Overall, the project demonstrated the effectiveness of time series analysis and forecasting techniques in understanding and predicting bike rental demand. Future work could include incorporating additional external factors such as detailed weather conditions or special events to further improve the model's accuracy.